General constraints on magic state protocols from the statistical mechanics of \\Wigner negativity

### 2021-06-10 magic_paper.tex
distributed refs. Remains to smooth narrative from intro to outlook + decide on figures + eliminate comments

\ddd{[Title of Gour2020 paper is incorrect in the bibliography]}

The past few years have seen rapid progress towards the goal of a fault-tolerant quantum computer~\cite{Nickerson_2014, Nikahd_2017, chao_2018, lin_pieceable_2020, Lin_2020, Bourassa_2021}. However, many challenges remain and there is increasing need for theory to take into account physical limitations of the hardware involved. The surface code~\cite{Bravyi_1998, Freedman_2001, Dennis_2002, Raussendorf_2007} is a leading framework for fault-tolerance with very high error thresholds. Within this scheme, Clifford unitaries can be implemented in a robust, fault-tolerant way. However, due to the Eastin-Knill theorem~\cite{Eastin_2009}, we also know that it is impossible to have a universal set of transversal gates, and although Clifford unitaries can be realised transversally~\cite{Calderbank_1996, Steane_1996}, one needs to find ways around the Eastin-Knill restriction. This can be achieved by injecting in quantum states, called magic states, which promote the Clifford group to universal quantum computing~\cite{cit:bravyi}. The obstacle to this is that these states are invariably noisy and so protocols involving stabilizer operations must be employed to purify many copies of the magic states and improve the overall performance of the induced quantum gates. A central question then arises about the overhead on purifying many copies of a magic state into less noisy forms.

To address this, concrete distillation protocols have been developed, such as the Bravyi-Haah protocol that provides a quadratic reduction in noise per cycle~\cite{Bravyi2012}. Such distillation rates have been improved in more recent works~\cite{Hastings2018, Litinski_2019, Krishna2019, cit:prakash}. There is also analysis of magic protocols from the perspective of magic monotones, which provide upper bounds on distillation rates and find application in the analysis of gate synthesis~\cite{Campbell_2017, Howard_2017, Prakash_2018}. These frameworks for magic view magic states as resource states with respect to a natural class of quantum operations that are considered cheap, or free, such as stabilizer operations~\cite{Gour_2019, cit:ahmadi, cit:seddon, Wang_2019}. Beyond quantum computing, recent work has also looked at how magic can be used to analyse many-body physics~\cite{Sarkar_2020} and conformal field theories~\cite{White_2021}.

Recent work towards fault-tolerance has begun to bridge the gap between abstract theory and experiment. Extensive work has been done on error mitigation~\cite{Li_2017, Temme_2017, Endo_2018, McClean_2017}, and the incorporation of hardware physics into the theoretical modes~\cite{Kandala_2019, Colless_2018, song2018quantum, Bravyi_2021}. For example, the XZZX code~\cite{bonilla_ataides_xzzx_2021} is a variant of the surface code that incorporates noise bias explicitly and has been shown to attain the hashing bound of random codes~\cite{Bennett_1996}.

@@

Through this approach we obtain a range of upper bounds for magic distillation rates. We provide explicit bounds for magic protocols that generate unital channels, as well as bounds that depend on the temperature and Hamiltonian of the system. We find that the thermodynamic ordering in the presence of negativity is non-trivial and displays a range of features that, unsurprisingly, do not appear in classical statistical mechanics. 
Finally, we discuss how our approach could be improved and how it can be exploited to construct explicit lower bounds on distillation that take into account physical structure.

@@

Our analysis applies majorization to magic states at the level of the associated Wigner distributions. However, these are in general quasi-probability distributions, so it is important to check how majorization is computed for these cases and what differences quasi-distributions bring over genuine probability distributions~\cite{arvidsson_2020, halpern_2018}.

@@

This motivates defining the following \emph{Wigner divergence} $D_\infty(\rho, \tau) := D_\infty (W_\rho(\x) || W_\tau (\x))$ on all quantum states. In~\cref{app:lcst_technical}, we show that this divergence has the following natural properties that are derived via relative majorization.
\begin{theorem}\label{thm:Dinfty} 
	Let $\tau$ be in the interior of $\F$. Then $D_\infty(W_\rho || W_\tau)$ is well-defined for all $\rho$, and the following hold:
\begin{enumerate}
\item $D_\infty(W_\rho || W_\tau) \ge 0$ for all quantum states $\rho$.
\item  $D_\infty(W_\rho || W_\tau) = 0$ if and only if $\rho =\tau$.
\item $D_\infty(W_{\rho^{\otimes 2n}}|| W_{\tau^{\otimes 2n}}) = n D_\infty(W_{\rho^{\otimes 2}} ||W_{\tau^{\otimes 2}})$ for any $n \in \mathbb{N}$.
\item For any quantum state $\rho$, and any free operation $\E$ such that $\E(\tau)$ is in the interior of $\F$, we have $D_\infty (W_\rho || W_\tau) \ge D_\infty (W_{\E(\rho)}|| W_{\E(\tau)})$.
\end{enumerate}
\end{theorem}

@@

We also note that the Renyi entropy can be viewed as a $q$--deformation of the Shannon entropy~\cite{baez2011renyi}, and $H_\alpha(\p)$ can be related to the $q$--derivative of free energy. More precisely, it can be shown that
\begin{equation}
	H_{\frac{T}{x}} = \frac{1}{x-T} (F(x) - F(T)),
\end{equation}
\nick{Do you mean $H_{\frac{T}{x}}(\w)$? What is $F(x)$ for a number $x$ ?}
where $F(x)$ is an effective free energy at temperature $x$. If we attempt to push this to the present setting, the presence of Wigner negativity would correspond to a divergence in the first derivative of this effective free energy as $x \rightarrow T$. One might speculate if the presence of magic in Clifford circuits admits a sensible interpretation in terms of a zeroth-order phase transition. However since variation in $\alpha$ is no longer smooth ($\alpha$ is restricted to a dense set of rational values) technical obstacles exist and we do not pursue this speculative line here, but just note that other works exist in which computational transitions are considered in relation to phase transitions.

Every entropy is best understood as a relative entropy~\cite{Gour_2020} and so we now state the extension to Renyi divergences.
We define the $\alpha$--Renyi Wigner divergence $D_\alpha$ as
\begin{equation}
	D_\alpha(\w || \r) := \frac{1}{\alpha-1} \log \sum_i w_i^\alpha r_i^{1-\alpha},
\end{equation}
where $\w$ can be a quasi-distribution, $\r$ a probability distribution with positive components and $\alpha$ can take the same values as for $H_\alpha$.
In~\cref{app:lcst_technical}, we also show that the divergence satisfies the following natural properties for all $\alpha$ for which it is well-defined.

\begin{theorem}\label{thm:Da_props} 
	Let $\tau$ be in the interior of $\F$. 
	If $\alpha = 2a / (2b-1)$ for non-negative integers $a,b$ with $a \geq b$, then the $\alpha$-Renyi divergence $D_\alpha(W_\rho || W_\tau)$ is well-defined for all states $\rho$, and the following hold:
\begin{enumerate}
\item $D_\alpha(W_\rho || W_\tau) \ge 0$ for all quantum states $\rho$.
\item  $D_\alpha(W_\rho || W_\tau) = 0$ if and only if $\rho =\tau$.
\item $D_\alpha(W_{\rho^{\otimes n}} || W_{\tau^{\otimes n}}) = n D_\alpha(W_\rho || W_\tau)$ for any $n \in \mathbb{N}$.
\item $D_\alpha(W_\rho || W_\tau) \geq D_\alpha (W_{\E(\rho)} || W_{\E(\tau)})$ for any free operation $\E$ such that $\E(\tau)$ is in the interior of $\F$.
\end{enumerate}
\end{theorem}

>>>>

The past few years have seen rapid progress towards the goal of a fault-tolerant quantum computer~\cite{cit:raussendorf, Raussendorf_2013, Nickerson_2014, Nikahd_2017, chao_2018, lin_pieceable_2020, Lin_2020, Bourassa_2021}. However, many challenges remain and there is increasing need for theory to take into account physical limitations of the hardware involved. The surface code~\cite{Bravyi_1998, Freedman_2001, Dennis_2002, Raussendorf_2007} is a leading framework for fault-tolerance with very high error thresholds. Within this scheme, Clifford unitaries can be implemented in a robust, fault-tolerant way. However, due to the Eastin-Knill theorem~\cite{Eastin_2009}, we also know that it is impossible to have a universal set of transversal gates, and although Clifford unitaries can be realised transversally~\cite{Calderbank_1996, Steane_1996}, one needs to find ways around the Eastin-Knill restriction. This can be achieved by injecting in quantum states, called magic states, which promote the Clifford group to universal quantum computing~\cite{cit:bravyi}. The obstacle to this is that these states are invariably noisy and so protocols involving stabilizer operations must be employed to purify many copies of the magic states and improve the overall performance of the induced quantum gates~\cite{cit:bravyi, Jones_2013, Ogorman_2017}. A central question then arises about the overhead on purifying many copies of a magic state into less noisy forms.

To address this, concrete distillation protocols have been developed, such as the Bravyi-Haah qubit protocol that provides a quadratic reduction in noise per cycle~\cite{Bravyi2012}. Such distillation rates have been improved in more recent works~\cite{Jones_2013, haah2017magic, Hastings2018, Litinski_2019} and protocols for qudits of odd dimension have also been proposed~\cite{Anwar_2012, CampbellAnwar_2012, Dawkins_2015, Krishna2019, cit:prakash}. There is also analysis of magic protocols from the perspective of magic monotones, which provide upper bounds on distillation rates and find application in the analysis of gate synthesis~\cite{Campbell_2017, Howard_2017, Prakash_2018, Seddon_2021}. These frameworks for magic view magic states as resource states with respect to a natural class of quantum operations that are considered cheap, or free, such as stabilizer operations~\cite{Gour_2019, cit:ahmadi, cit:seddon, Wang_2019}. Beyond quantum computing, recent work has also looked at how magic can be used to analyse many-body physics~\cite{Sarkar_2020} and conformal field theories~\cite{White_2021} as well as how it is connected with contextuality~\cite{Vega_2017, cit:howard2, Zurel_2020, okay2021extremal}.

Recent work towards fault-tolerance has begun to bridge the gap between abstract theory and experiment. Extensive work has been done on error mitigation~\cite{Li_2017, Temme_2017, Endo_2018, McClean_2017}, and the incorporation of hardware physics into the theoretical modes~\cite{holmes_resource_2019, Kandala_2019, Colless_2018, song2018quantum, Bravyi_2021}. For example, the XZZX code~\cite{bonilla_ataides_xzzx_2021} is a variant of the surface code that incorporates noise bias explicitly and has been shown to attain the hashing bound of random codes~\cite{Bennett_1996}.

@@

Through this approach, we obtain a range of upper bounds for magic distillation rates. We provide explicit bounds for magic protocols that generate unital channels, as well as bounds that depend on the temperature and Hamiltonian of the system. We find that the thermodynamic ordering in the presence of negativity is non-trivial and displays a range of features that, unsurprisingly, do not appear in classical statistical mechanics. 
Finally, we discuss how our approach could be improved and how it can be exploited to construct explicit lower bounds on distillation that take into account physical structure.

@@

Our analysis applies majorization to magic states at the level of the associated Wigner distributions. However, these are in general quasi-probability distributions, so it is important to check how majorization is computed for these cases and what differences quasi-distributions bring over genuine probability distributions.

@@

This motivates defining the following \emph{Wigner divergence} $D_\infty (W_\rho(\x) || W_\tau (\x))$ on all quantum states. In~\cref{app:lcst_technical}, we show that this divergence has the following natural properties that are derived via relative majorization.
\begin{theorem}\label{thm:Dinfty} 
	Let $\tau$ be in the interior of $\F$. Then $D_\infty(W_\rho || W_\tau)$ is well-defined for all $\rho$, and the following hold:
\begin{enumerate}
\item $D_\infty(W_\rho || W_\tau) \ge 0$ for all quantum states $\rho$.
\item  $D_\infty(W_\rho || W_\tau) = 0$ if and only if $\rho =\tau$.
\item $D_\infty(W_{\rho^{\otimes 2n}}|| W_{\tau^{\otimes 2n}}) = n D_\infty(W_{\rho^{\otimes 2}} ||W_{\tau^{\otimes 2}})$ for any $n \in \mathbb{N}$.
\item $D_\infty(W_\rho || W_\tau) \geq D_\infty(W_{\E(\rho)} || W_{\E(\tau)})$ for any free operation $\E$ such that $\E(\tau)$ is in the interior of $\F$.
\end{enumerate}
\end{theorem}

@@

We also note that the Renyi entropy can be viewed as a $q$--deformation of the Shannon entropy~\cite{baez2011renyi}, and $H_\alpha(\p)$ can be related to the $q$--derivative of free energy. More precisely, it can be shown that
\begin{equation}
	H_{\frac{T}{x}} = \frac{1}{x-T} (F(x) - F(T)),
\end{equation}
\nick{Do you mean $H_{\frac{T}{x}}(\w)$? What is $F(x)$ for a number $x$ ?}
where $F(x)$ is an effective free energy at temperature $x$. If we attempt to push this to the present setting, the presence of Wigner negativity would correspond to a divergence in the first derivative of this effective free energy as $x \rightarrow T$. One might speculate if the presence of magic in Clifford circuits admits a sensible interpretation in terms of a zeroth-order phase transition. However since variation in $\alpha$ is no longer smooth ($\alpha$ is restricted to a dense set of rational values) technical obstacles exist and we do not pursue this speculative line here, but just note that other works exist in which computational transitions are considered in relation to phase transitions~\cite{Allahverdyan_2018, arvidsson_2020, halpern_2018, Lostaglio_2018, Levy_2020} \nick{I don't understand this statement, so I just dropped all citations on quasi-distributions here}.

Every entropy is best understood as a relative entropy~\cite{Gour_2020} and so we now state the extension to Renyi divergences.
We define the $\alpha$--Renyi Wigner divergence $D_\alpha(\p||\r)$ as
\begin{equation}
	D_\alpha(\p || \r) := \frac{1}{\alpha-1} \log \sum_i p_i^\alpha r_i^{1-\alpha},
\end{equation}
where $\p$ can be a quasi-distribution, $\r$ a probability distribution with positive components and $\alpha$ can take the same values as for $H_\alpha$.
In~\cref{app:lcst_technical}, we also show that the divergence satisfies the following natural properties for all $\alpha$ for which it is well-defined.

\begin{theorem}\label{thm:Da_props} 
	Let $\tau$ be in the interior of $\F$. 
	If $\alpha = \frac{2a}{2b-1}$ for non-negative integers $a,b$ with $a \geq b$, then the $\alpha$-Renyi divergence $D_\alpha(W_\rho || W_\tau)$ is well-defined for all states $\rho$, and the following hold:
\begin{enumerate}
\item $D_\alpha(W_\rho || W_\tau) \ge 0$ for all quantum states $\rho$.
\item  $D_\alpha(W_\rho || W_\tau) = 0$ if and only if $\rho =\tau$.
\item $D_\alpha(W_{\rho^{\otimes n}} || W_{\tau^{\otimes n}}) = n D_\alpha(W_\rho || W_\tau)$ for any $n \in \mathbb{N}$.
\item $D_\alpha(W_\rho || W_\tau) \geq D_\alpha (W_{\E(\rho)} || W_{\E(\tau)})$ for any free operation $\E$ such that $\E(\tau)$ is in the interior of $\F$.
\end{enumerate}
\end{theorem}
